First Order Alternation

### 2018-10-16 quantifiers/draft.tex
ok

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X \subseteq \vars$, we
denote by $X \mapsto \Data$ the set of valuations of the variables $X$
and $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly
infinite set of \emph{data symbols} $(a,\nu)$, where $a$ is an input
symbol and $\nu$ is a valuation. A \emph{data word} (simply called
word in the following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2)
\ldots (a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$. A
\emph{first-order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
\item $\Sigma$ is a finite set of input events, 
\item $X$ is a finite set of input variables, 
\item $Q$ is a finite set of predicates denoting control states, 
\item $\iota \in \posforms(Q,\emptyset)$ is a sentence defining initial configurations,
\item $F \subseteq Q$ is the set of predicates denoting final states, and 
\item $\Delta$ is a set of \emph{transition rules} of the form
  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is
  a predicate, $a \in \Sigma$ is an input event and $\psi \in
  \posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive
  formula, where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.
\end{compactitem}
The quantifiers occurring in the right-hand side formula of a
transition rule are referred to as \emph{transition quantifiers}. The
\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

The intuition of a transition rule \(q(y_1,\ldots,y_{\#(q)})
\arrow{a(X)}{} \psi\) is the following: $a$ is the input event and $X$
are the input data values that trigger the transition, whereas $q$ and
$y_1,\ldots,y_{\#(q)}$ are the current control state and data values
in that state, respectively. Without loss of generality, we consider,
for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
most one such rule --- two or more rules can be joined using
disjunction.

The execution semantics of automata is given in close analogy with the
case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
is a model but is not minimal\footnote{To avoid the confusion of
  considering potentially non-minimal models, the original definition
  of alternating finite-state automata \cite{ChandraKozenStockmeyer81}
  uses boolean valuations instead of formulae.}.

@@

Typically, when checking the unreachability of a set of program
configurations or, more generally, the infeasability of a set of Horn
clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}, the interpolants
used to annotate the unfolded control structure are assertions about
the values of the program variables in a given control state, at a
certain step of a given execution. However, in an alternating model of
computation, it is useful to distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

\paragraph{Remark}
When considering an alternating model of computation, control-local
interpolants are not always enough to prove emptiness, because of the
synchronization of several branches of the computation on the same
sequence of input values. Consider, for instance, an automaton with
the following transition rules and final state $q_f$:
\[\begin{array}{rclcrcl}
q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
q_1(y) \arrow{a(x)}{} y+x > 0 \wedge q_f \\
q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \wedge q_f \\
q_2(y) & \arrow{a(x)}{} & q_2(y-x)
\end{array}\]
Started in an initial configuration $q_0(0)$ with an input word
$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
the automaton executes as follows:
\[\begin{array}{l}
q_0(0) \arrow{(a,\nu_1)}{} q_1(k_1)
\wedge q_2(-k_1) \ldots \arrow{(a,\nu_{n-1})}{} 
q_1(\sum_{i=1}^{n-1} k_i) \wedge q_2(-\sum_{i=1}^{n-1} k_i) \arrow{(a,\nu_n)}{} \\
\sum_{i=1}^{n} k_i > 0 \wedge \sum_{i=1}^{n} k_i < 0 \wedge q_f \equiv \bot
\end{array}\] 
An overapproximation of the set of cubes generated after one or more
steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a control-local
formula using one occurrence of a predicate would give a too rough
overapproximation of this set, unable to prove the emptiness of the
automaton. \hfill$\blacksquare$

@@

\begin{definition}[\cite{Lyndon59}]\label{def:lyndon-interpolant}
Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
unsatisfiable, a \emph{Lyndon interpolant} is a formula $I$ such that
$\phi \models I$, the formula $I \wedge \psi$ is unsatisfiable,
$\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, $\pset{+}{I} \subseteq
\pset{+}{\phi} \cap \pset{+}{\psi}$ and $\pset{-}{I} \subseteq
\pset{-}{\phi} \cap \pset{-}{\psi}$.
\end{definition}
The existence of Lyndon interpolants for first-order logic, with no
predefined interpretation of function symbols\footnote{A model
  specifies the universe and the interpretation of function symbols.},
was proved by a non-constructive model-theoretic argument
\cite{Lyndon59}. In the following, we shall consider only those
theories of the data domain $\Data$, for which there exists a decision
procedure for satisfiability, able to produce Lyndon interpolants. For
instance, such interpolating decision procedures exist for linear real
and integer arithmetic, with uninterpreted function symbols (UFLRA and
UFLIA) \cite{RybalchenkoSofronieStokkermans}. In the rest of this
section, let us fix an automaton $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above observation, none
of the interpolants considered will be control-local and we shall use
the term \emph{local} to denote time-local interpolants, with no free
variables.

@@

The main problem with the local GLI construction described in the
proof of Proposition \ref{prop:local-gli} is that the existence of
Lyndon interpolants (Definition \ref{def:lyndon-interpolant}) is
guaranteed in principle, but the proof is non-constructive. Building
an interpolant for an unsatisfiable conjunction of formulae $\phi
\wedge \psi$ is typically the job of the decision procedure that
proves the unsatisfiability and, in general, there is no such
procedure, when $\phi$ and $\psi$ contain predicates and have
non-trivial quantifier alternation. In this case, some provers use
instantiation heuristics for the universal quantifiers that are
sufficient for proving unsatisfiability, however these heuristics are
not always suitable for interpolant generation. Consequently, from now
on, we assume the existence of an effective Lyndon interpolation
procedure only for decidable theories, such as the quantifier-free
linear (integer) arithmetic with uninterpreted functions (QFLIA,
QFLRA, etc.).

@@

The following lemma gives the formal proof of correctness for this
construction.

@@

    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where $\set{i_1, \ldots,
    i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}$ and
    $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k
    \mid k < i_j, Q_k = \exists}$, for all $j \in [1,\ell]$, such that

@@

    \ldots, I_n)$ for $\alpha$, such that $\voc{I_k} \subseteq

@@

      \max(\xi^{-1}(k)),~ Q_j = \exists}$, for all $k \in [n]$.

@@

  $\set{i_1,\ldots,i_\ell}=\set{i \in [1,m] \mid Q_i=\forall}$, such
  that: \begin{compactitem}

@@

    of some prenex form of the formula:

@@

  form of the formula:

@@

  \noindent(\ref{it2:subst-quant}) With the notation introduced at
  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
  \tau_{i_\ell}$, whose existence is proved by point
  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi$ is
  monotonic, we obtain: \begin{compactitem}
  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
    \stamp{X}{0} \cup \set{x_j \mid j < \max(\xi^{-1}(0)), Q_j =
      \exists}$,
    %
  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
      \max(\xi^{-1}(i)), Q_j = \exists}$, for all $i \in [1,n]$,
    %
  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
  \end{compactitem}
  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
        \max(\xi^{-1}(0)), Q_j = \exists}$,
      %
    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
        \max(\xi^{-1}(k)), Q_j = \exists}$, for all $k \in [1,n]$,
      %
    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
  \end{compactitem}
  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
      Theorem \ref{thm:skolem},
      %
    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
      \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
      I_{k-1} \wedge \phi_k \models I_k$, and 
      %
    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
      \models I_n \wedge \psi \models \bot$,     
  \end{compactitem}
  as required by Definition
  \ref{def:generalized-lyndon-interpolant}. \qed}

>>>>

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X \subseteq \vars$, we
denote by $X \mapsto \Data$ the set of valuations of the variables $X$
and $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly
infinite set of \emph{data symbols} $(a,\nu)$, where $a$ is an input
symbol and $\nu$ is a valuation. A \emph{data word} (simply called
word in the following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2)
\ldots (a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$.

Formally, a \emph{first-order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
occurring in the right-hand side formula of a transition rule are
referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

The intuition of a transition rule \(q(y_1,\ldots,y_{\#(q)})
\arrow{a(X)}{} \psi\) is the following: $a$ is the input event and $X$
are the input data values that trigger the transition, whereas $q$ and
$y_1,\ldots,y_{\#(q)}$ are the current control state and data values
in that state, respectively. Without loss of generality, we consider,
for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
most one such rule, as two or more rules can be joined using
disjunction.

The execution semantics of automata is given in close analogy with the
case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
is a model but is not minimal. The original definition of alternating
finite-state automata \cite{ChandraKozenStockmeyer81} works around
this problem by considering boolean valuations (models) instead of
formulae.

@@

Typically, when checking the unreachability of a set of program
configurations \cite{mcmillan06}, the interpolants used to annotate
the unfolded control structure are assertions about the values of the
program variables in a given control state, at a certain step of a
given execution. However, in an alternating model of computation, it
is useful to distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

\paragraph{Remark}
When considering an alternating model of computation, control-local
interpolants are not always enough to prove emptiness, because of the
synchronization of several branches of the computation on the same
sequence of input values. Consider, for instance, an automaton with
the following transition rules and final state $q_f$:
\[\begin{array}{rclcrcl}
q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
q_1(y) \arrow{a(x)}{} y+x > 0 \wedge q_f \\
q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \wedge q_f \\
q_2(y) & \arrow{a(x)}{} & q_2(y-x)
\end{array}\]
Started in an initial configuration $q_0(0)$ with an input word
$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
the automaton executes as follows:
\[\begin{array}{c}
q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots \arrow{(a,\nu_{n-1})}{} 
\set{q_1(\sum_{i=1}^{n-1} k_i),q_2(-\sum_{i=1}^{n-1} k_i)}
\arrow{(a,\nu_n)}{} \emptyset
\end{array}\]
An overapproximation of the set of cubes generated after one or more
steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a control-local
formula using one occurrence of a predicate would give a too rough
overapproximation of this set, unable to prove the emptiness of the
automaton. \hfill$\blacksquare$

@@

\begin{definition}[\cite{Lyndon59}]\label{def:lyndon-interpolant}
Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
unsatisfiable, a \emph{Lyndon interpolant} is a formula $I$ such that
$\phi \models I$, the formula $I \wedge \psi$ is unsatisfiable,
$\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, $\pset{+}{I} \subseteq
\pset{+}{\phi} \cap \pset{+}{\psi}$ and $\pset{-}{I} \subseteq
\pset{-}{\phi} \cap \pset{-}{\psi}$.
\end{definition}
%% The existence of Lyndon interpolants for first-order logic, with no
%% predefined interpretation of function symbols\footnote{A model
%%   specifies the universe and the interpretation of function symbols.},
%% was proved by a non-constructive model-theoretic argument
%% \cite{Lyndon59}. In the following, we shall consider only those
%% theories of the data domain $\Data$, for which there exists a decision
%% procedure for satisfiability, able to produce Lyndon interpolants. For
%% instance, such interpolating decision procedures exist for linear real
%% and integer arithmetic, with uninterpreted function symbols (UFLRA and
%% UFLIA).

In the rest of this section, let us fix an automaton $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above observation, none
of the interpolants considered will be control-local and we shall use
the term \emph{local} to denote time-local interpolants, with no free
variables.

@@

The main problem with the local GLI construction described in the
proof of Proposition \ref{prop:local-gli} is that the existence of
Lyndon interpolants (Definition \ref{def:lyndon-interpolant}) is
guaranteed in principle, but the proof is non-constructive. Building
an interpolant for an unsatisfiable conjunction of formulae $\phi
\wedge \psi$ is typically the job of the decision procedure that
proves the unsatisfiability and, in general, there is no such
procedure, when $\phi$ and $\psi$ contain predicates and have
non-trivial quantifier alternation. In this case, some provers use
instantiation heuristics for the universal quantifiers that are
sufficient for proving unsatisfiability, however these heuristics are
not always suitable for interpolant generation. Consequently, from now
on, we assume the existence of an effective Lyndon interpolation
procedure only for decidable theories, such as the quantifier-free
linear (integer) arithmetic with uninterpreted functions (UFLIA,
UFLRA, etc.) \cite{RybalchenkoSofronieStokkermans}.

@@

A function $\xi : \nat \rightarrow \nat$ is [strictly]
\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$
     [$\xi(n) < \xi(m)$] and \emph{finite-range} iff for each $n \in
     \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
     finite-range, we denote by $\xi^{-1}(n) \in \nat$ the maximal
     value $m$ such that $\xi(m)=n$. It is easy to see that $\xi^{-1}$
     is strictly monotonic if $\xi$ is monotonic and finite-range. The
     lemma below gives the proof of correctness for the previous
     construction of non-local GLI.

@@

    \[\begin{array}{l}   
    \set{i_1, \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall} \text{ and} \\
    \voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k
    \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]
    \end{array}\]
    such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,

@@

    \ldots, I_n)$ for $\alpha$, such that \[\voc{I_k} \subseteq

@@

      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\enspace.\]

@@

  $\set{i_1,\ldots,i_\ell}=\set{i \in [1,m] \mid Q_i=\forall}$
  and: \begin{compactitem}

@@

    of some prenex form of:

@@

  form of:

@@

  \noindent(\ref{it2:subst-quant}) With the notation introduced at
  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
  \tau_{i_\ell}$, whose existence is proved by point
  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi^{-1}$ is
  strictly monotonic, we obtain: \begin{compactitem}
  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
    \stamp{X}{0} \cup \set{x_j \mid j < \xi^{-1}(0), Q_j = \exists}$,
    %
  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
      \max(\xi^{-1}(i)), Q_j = \exists}$, for all $i \in [1,n]$,
    %
  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
  \end{compactitem}
  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
        \max(\xi^{-1}(0)), Q_j = \exists}$,
      %
    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
        \max(\xi^{-1}(k)), Q_j = \exists}$, for all $k \in [1,n]$,
      %
    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
  \end{compactitem}
  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
      Theorem \ref{thm:skolem},
      %
    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
      \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
      I_{k-1} \wedge \phi_k \models I_k$, and 
      %
    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
      \models I_n \wedge \psi \models \bot$,     
  \end{compactitem}
  as required by Definition
  \ref{def:generalized-lyndon-interpolant}. \qed}
